---
title: "Predictors of Premature Mortality at the County Level in the United States"
subtitle: "County-Level Public Health Analysis"
author: ""
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{skip=4pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.pos = "H", out.width = "80%"
)
suppressPackageStartupMessages({
  library(readxl)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(knitr)
})
theme_set(theme_minimal(base_size = 10))
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)
```

```{r load-merge-clean}
chr_path <- "data/county_health_rankings_2018.xls"
op_path  <- "data/opioid_prescribing_2016.csv"

# Dependent variable: Premature age-adjusted mortality (Sheet 6, "Age-Adjusted Mortality")
dv <- read_excel(chr_path, sheet = 6, skip = 2, col_names = FALSE)[, c(1, 5)]
names(dv) <- c("FIPS", "premature_aamort")

# Internal predictors (Sheet 4, "Ranked Measure Data")
iv <- read_excel(chr_path, sheet = 4, skip = 2, col_names = FALSE)[, c(1, 31, 35, 68, 105, 117)]
names(iv) <- c("FIPS", "pct_smokers", "pct_obese", "pct_uninsured",
               "grad_rate", "pct_child_poverty")

# External predictor: CDC county opioid dispensing rate, 2016
op <- read_csv(op_path, show_col_types = FALSE)[, c(3, 4)]
names(op) <- c("FIPS", "opioid_rate_2016")

county <- dv %>%
  left_join(iv, by = "FIPS") %>%
  mutate(FIPS = as.numeric(FIPS)) %>%
  left_join(mutate(op, FIPS = as.numeric(FIPS)), by = "FIPS") %>%
  filter(!is.na(FIPS), FIPS > 1000, FIPS %% 1000 != 0) %>%   # counties only
  mutate(across(-FIPS, ~ suppressWarnings(as.numeric(.))))

vars <- c("premature_aamort", "pct_smokers", "pct_obese",
          "pct_uninsured", "grad_rate", "pct_child_poverty", "opioid_rate_2016")
labels <- c(
  premature_aamort  = "Premature age-adjusted mortality (per 100,000)",
  pct_smokers       = "Adult smoking (%)",
  pct_obese         = "Adult obesity (%)",
  pct_uninsured     = "Uninsured (%)",
  grad_rate         = "High-school graduation rate (%)",
  pct_child_poverty = "Children in poverty (%)",
  opioid_rate_2016  = "Opioid dispensing rate, 2016 (per 100)"
)

analysis <- county %>% select(all_of(vars)) %>% tidyr::drop_na()
n_total <- nrow(county); n_complete <- nrow(analysis)
```

\newpage

# Introduction

County-level health outcomes vary widely across the United States, and the
factors behind that variation matter for how scarce public-health resources are
allocated. This report answers the question: **what predicts better health
outcomes at the county level in the United States?** Following the project
brief, the outcome is *premature age-adjusted mortality* (deaths before age 75,
age-standardised, per 100,000 population) from the 2018 County Health Rankings.
Because lower mortality represents better health, the aim is to identify which
county-level social, behavioural, healthcare-access, and drug-environment
variables are most strongly associated with better health outcomes.

Premature mortality is a useful summary of population health because it weights
the deaths that are, in principle, most avoidable through prevention and timely
care. Measuring it by county matters for policy, because much U.S. health
funding and programme delivery is administered locally, so identifying which
county characteristics travel with worse outcomes helps target limited
resources. The variation is also large enough to be worth explaining: across the
analysis sample, premature mortality ranges from about
`r fmt(min(analysis$premature_aamort),0)` to
`r fmt(max(analysis$premature_aamort),0)` deaths per 100,000.

This study argues that premature mortality is shaped jointly by four domains:
health behaviour, socioeconomic disadvantage, healthcare access, and the local
drug environment. It tests this by building a multivariate linear regression
model with six predictors and one externally sourced variable (CDC opioid
dispensing). Previewing the main findings: socioeconomic disadvantage (child
poverty) and behavioural risk (smoking) are the strongest predictors, opioid
dispensing adds independent explanatory power, and — contrary to expectation —
graduation rate is *not* protective once the other variables are controlled,
a pattern this report interprets as a confounding/suppression effect.

# Hypotheses

- **General hypothesis:** counties with healthier behavioural profiles,
  lower socioeconomic disadvantage, better healthcare access, and a less harmful
  drug environment have lower premature mortality.

The specific hypotheses below state both the expected *direction* and, drawing on
the literature reviewed in the next section, the expected *strength* of each
association.

- **H1 (smoking):** higher adult smoking rate → higher premature mortality;
  expected to be one of the strongest predictors, as smoking is the leading
  preventable cause of death.
- **H2 (obesity):** higher adult obesity rate → higher premature mortality;
  expected to be a moderate predictor, since the obesity–mortality association is
  smaller and more contested than that for smoking.
- **H3 (uninsured):** higher uninsured rate → higher premature mortality;
  expected to be a moderate predictor.
- **H4 (education):** higher high-school graduation rate → *lower* premature
  mortality; expected to be protective but largely *indirect*, so its independent
  effect may be weak once behaviour and poverty are held constant.
- **H5 (child poverty):** higher child poverty rate → higher premature mortality;
  expected to be among the strongest predictors, as socioeconomic disadvantage is
  a foundational determinant of health.
- **H6 (opioids):** higher opioid dispensing rate → higher premature mortality;
  expected to be positive but secondary to the socioeconomic and behavioural
  factors.

# Literature and theory

County premature mortality varies widely across the United States, and that
variation tracks social conditions far more than chance. This study draws on the
*social determinants of health* framework, which links health to the
circumstances in which people live, work, and age. In a multicohort study of 1.7 million
people, Stringhini et al. (2017) show that low socioeconomic status predicts
premature mortality about as strongly as established behavioural risk factors,
and that these disadvantages accumulate over the life course. Building on this
framework, the present study treats county premature mortality as jointly shaped
by four established domains — socioeconomic disadvantage, educational
attainment, healthcare access, and behavioural risk — with a fifth contextual
factor, the local drug environment, added for this project.

Socioeconomic disadvantage is the foundation. Childhood poverty is
associated with worse health across the life course and elevated mortality risk
(Wise, 2016), in part through early-life biological embedding that raises later
cardiovascular risk (Galobardes et al., 2006). Counties with higher child
poverty are therefore expected to show higher premature mortality (H5,
positive).

Education is consistently one of the strongest predictors of longevity. A global
systematic review finds that additional schooling lowers adult mortality
(IHME-CHAIN Collaborators, 2024), a gradient confirmed in U.S. cohorts (Kaplan
et al., 2015). Crucially, much of education's protective effect operates
*indirectly*, through the behaviours and material conditions it shapes: Puka et
al. (2022) show that lifestyle risk factors such as smoking explain a substantial
share of the education–mortality gradient. This mediated structure is central to
the analysis, because it implies that the graduation-rate coefficient may weaken
or even change sign once smoking and poverty are held constant (H4, expected
negative but plausibly mediated).

Access to care matters independently of behaviour. Systematic reviews show that health
insurance increases the use of effective care and improves outcomes (Freeman et
al., 2008), and that uninsured U.S. adults face higher mortality (Wilper et al.,
2009; Woolhandler & Himmelstein, 2017). This evidence is weaker than it first
appears, however: it is largely observational and open to reverse causation and
selection — sicker and poorer people are also more likely to be uninsured — so the
independent effect of insurance is harder to isolate than that of a behavioural
risk. A higher uninsured rate is therefore expected to predict higher mortality,
though plausibly only modestly (H3, positive).

Behaviour is the most direct pathway of all. Smoking remains the leading preventable
cause of death in the United States (U.S. Department of Health and Human
Services, 2014), is strongly associated with all-cause mortality (Lariscy et
al., 2018), and its harms are at least partly reversible after cessation (Mons et
al., 2015). Obesity likewise raises all-cause mortality risk across large
meta-analyses (Flegal et al., 2013; Aune et al., 2016; Jayedi et al., 2022),
though the size of this association is contested: estimates vary with how body
mass is categorised, and BMI is an imperfect proxy for adiposity (Flegal et al.,
2013). Both smoking and obesity are therefore expected to be positive predictors,
with obesity the weaker of the two (H1, H2).

Finally, this project adds the local drug environment. County-level drug
mortality is closely tied to the supply of prescription opioids and to the
economic distress of place (Monnat, 2018; Dasgupta et al., 2018). This evidence is
itself largely ecological — established across places rather than individuals — so it
justifies including the opioid dispensing rate as an external predictor without, on
its own, establishing individual-level risk (H6, positive).

These domains do not act in isolation. Most of education's protective effect is
expected to run through behaviour and material conditions rather than directly,
which is why the graduation-rate coefficient may weaken once those factors are
held constant. Because these determinants are measured at the
county rather than the individual level, the analysis speaks to ecological
association rather than individual causation, and aggregate correlations need not
hold within individuals (Robinson, 1950).

# Data and methodology

## Data sources

```{r data-source-table}
src <- data.frame(
  Dataset = c("County Health Rankings 2018", "CDC opioid dispensing, 2016"),
  Provider = c("Robert Wood Johnson Foundation & Univ. of Wisconsin Population Health Institute",
               "U.S. Centers for Disease Control and Prevention"),
  Role = c("Outcome + 5 predictors", "External predictor (opioid_rate_2016)"),
  Merge = c("FIPS county code", "FIPS county code")
)
kable(src, booktabs = TRUE, align = c("l", "l", "l", "l"),
      caption = "Data sources and their role in the analysis.")
```

Table 1 summarises the two datasets and their roles. The provided dataset is
the 2018 County Health Rankings (RWJF & University of Wisconsin Population Health
Institute, 2018). The outcome, *premature
age-adjusted mortality*, is taken from the Additional Measure Data sheet. Five
predictors — adult smoking, adult obesity, uninsured rate, high-school
graduation rate, and children in poverty — are drawn from the Ranked Measure
Data sheet. As required, an additional external dataset is merged in: the CDC
county-level opioid dispensing rate for 2016 (Centers for Disease Control and
Prevention, 2017), joined on the five-digit FIPS county code.

The County Health Rankings are reasonably reliable because their component
measures draw from established sources such as the CDC, the Census Bureau's
American Community Survey, and the National Center for Education Statistics.
Some measures are modelled estimates with confidence intervals, so county values
carry sampling uncertainty. Bias mitigation is limited but explicit: counties
are merged by FIPS rather than names, state roll-ups are filtered out, missingness
is reported before modelling, and the 2016 opioid rate is treated as a lagged
exposure rather than a contemporaneous measure.

## Variables and descriptive statistics

```{r descriptives}
desc <- sapply(analysis, function(x) c(
  N = sum(!is.na(x)), Mean = mean(x), SD = sd(x),
  Min = min(x), Median = median(x), Max = max(x)))
desc <- as.data.frame(t(round(desc, 1)))
desc <- cbind(Variable = labels[rownames(desc)], desc)
rownames(desc) <- NULL
kable(desc, booktabs = TRUE, row.names = FALSE,
      align = c("l", rep("r", ncol(desc) - 1)),
      caption = "Descriptive statistics, complete-case analysis sample.")
```

Table 2 reports the mean, spread, and range of each variable across the
complete-case sample; the predictors are measured on comparable percentage or
rate scales, while the opioid dispensing rate shows the widest dispersion.

## Missing data

```{r missingness}
miss <- county %>% select(all_of(vars)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing")
miss$Variable <- labels[miss$Variable]
kable(miss, booktabs = TRUE, align = c("l", "r"),
      caption = sprintf("Missing values by variable (of %d counties). Complete-case N = %d.",
                        n_total, n_complete))
```

Listwise deletion is used, giving a complete-case sample of `r n_complete`
counties out of `r n_total`. As Table 3 shows, most missingness is driven by the
graduation rate (`r sum(is.na(county$grad_rate))` counties) and the opioid
dispensing rate (`r sum(is.na(county$opioid_rate_2016))` counties).

Listwise deletion is a pragmatic but imperfect choice: it gives a transparent
regression sample, but may under-represent smaller or more rural counties because
missingness is concentrated in graduation and opioid data. Imputation is not used
because assuming those counties resemble reporting counties is untestable; this
representativeness issue is revisited in the limitations.

## Method

No log transformation or recoding is applied in the main model because the
selected variables are already interpretable percentages or rates. Coefficients
are kept in their original units for substantive interpretation, while
standardised coefficients are used only as a supplementary way to compare
relative predictor strength.

The outcome is continuous, so the analysis uses **multiple linear regression
(ordinary least squares)** with the six predictors entered simultaneously. OLS
is chosen because it directly estimates the partial association of each
predictor with mortality, holding the others constant, and the coefficients are
interpretable as a change in deaths per 100,000. In the full model, each
predictor also acts as a control variable for the others. The model is:

$$
\text{Mortality}_i = \beta_0 + \beta_1 \text{Smoking}_i + \beta_2 \text{Obesity}_i
+ \beta_3 \text{Uninsured}_i + \beta_4 \text{Graduation}_i
+ \beta_5 \text{ChildPoverty}_i + \beta_6 \text{Opioid}_i + \varepsilon_i
$$

Three modelling assumptions are examined below: linearity and constant error
variance through a residuals-versus-fitted plot, and multicollinearity through
variance inflation factors. As a method, OLS is well suited to this task — it
yields transparent, directly interpretable partial coefficients — but it also
assumes a linear, additive specification with independent errors, so it cannot
capture spatial dependence between neighbouring counties or any non-linear
effects. One feature of the design cannot be addressed by
diagnostics. Because every variable is measured for the county as a whole, the
coefficients describe differences between places rather than between people, and
reading a county-level association as an individual-level effect would be an
ecological fallacy; the interpretation throughout is framed accordingly.

# Results

## Exploratory data analysis

```{r distributions, fig.height=4.6, out.width="100%", fig.cap="Distributions of the outcome and predictors across counties."}
dist_labels <- c(premature_aamort = "Mortality", pct_smokers = "Smoking",
                 pct_obese = "Obesity", pct_uninsured = "Uninsured",
                 grad_rate = "Graduation", pct_child_poverty = "Child poverty",
                 opioid_rate_2016 = "Opioid")
analysis %>%
  rename_with(~ dist_labels[.x]) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 40, fill = "#2C6E91", colour = "white", linewidth = 0.1) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Counties")
```

```{r corr-heatmap, fig.height=3.8, fig.cap="Pearson correlations among the outcome and predictors."}
short <- c(premature_aamort = "Mortality", pct_smokers = "Smoking",
           pct_obese = "Obesity", pct_uninsured = "Uninsured",
           grad_rate = "Graduation", pct_child_poverty = "Child poverty",
           opioid_rate_2016 = "Opioid")
cmat <- cor(analysis)
dimnames(cmat) <- list(short[rownames(cmat)], short[colnames(cmat)])
cdf <- as.data.frame(as.table(cmat))
names(cdf) <- c("V1", "V2", "r")
ggplot(cdf, aes(V1, V2, fill = r)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = fmt(r, 2)), size = 2.7) +
  scale_fill_gradient2(low = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0, limits = c(-1, 1)) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(x = NULL, y = NULL, fill = "r")
```

The outcome ranges from `r fmt(min(analysis$premature_aamort),0)` to
`r fmt(max(analysis$premature_aamort),0)` premature deaths per 100,000
(mean `r fmt(mean(analysis$premature_aamort),0)`). Bivariately (Figure 2), child poverty
(r = `r fmt(cor(analysis$pct_child_poverty, analysis$premature_aamort))`) and
smoking (r = `r fmt(cor(analysis$pct_smokers, analysis$premature_aamort))`) are
the strongest correlates; graduation rate is only weakly (and negatively)
associated (r = `r fmt(cor(analysis$grad_rate, analysis$premature_aamort))`).

Two features of the data shape what follows. First, the predictors are
themselves correlated — adult smoking and obesity move together across counties
(r = `r fmt(cor(analysis$pct_smokers, analysis$pct_obese))`), as do poverty and
several behavioural measures — so the partial coefficients from the full model
can differ from these simple correlations once the variables are entered
together. Second, as Figure 1 shows, most variables are roughly symmetric, but
the opioid dispensing rate is right-skewed, with a long tail of
high-dispensing counties.
The strong bivariate links from child poverty and smoking to mortality already
foreshadow their prominence in the regression.

## Regression model

```{r model}
m <- lm(premature_aamort ~ pct_smokers + pct_obese + pct_uninsured +
          grad_rate + pct_child_poverty + opioid_rate_2016, data = analysis)
co <- summary(m)$coefficients
ci <- confint(m)
reg <- data.frame(
  Predictor = c("(Intercept)", labels[c("pct_smokers","pct_obese","pct_uninsured",
                "grad_rate","pct_child_poverty","opioid_rate_2016")]),
  b = fmt(co[, 1], 3),
  SE = fmt(co[, 2], 3),
  `95% CI` = sprintf("[%s, %s]", fmt(ci[, 1], 2), fmt(ci[, 2], 2)),
  t = fmt(co[, 3], 2),
  p = ifelse(co[, 4] < .001, "<.001", fmt(co[, 4], 3)),
  check.names = FALSE
)
kable(reg, booktabs = TRUE, row.names = FALSE,
      align = c("l", "r", "r", "r", "r", "r"),
      caption = sprintf("OLS regression predicting premature age-adjusted mortality. Coefficients are unstandardised; 95 per cent confidence intervals in brackets. N = %d; R-squared = %.3f; adjusted R-squared = %.3f.",
                        nobs(m), summary(m)$r.squared, summary(m)$adj.r.squared))
```

```{r vif, include=FALSE}
# VIF without the car package: for an OLS model with only continuous
# predictors, VIF_j = 1 / (1 - R^2_j), where R^2_j is obtained by
# regressing predictor j on all the other predictors.
X <- model.matrix(m)[, -1, drop = FALSE]
vif_values <- setNames(vapply(seq_len(ncol(X)), function(j)
  1 / (1 - summary(lm(X[, j] ~ X[, -j, drop = FALSE]))$r.squared),
  numeric(1)), colnames(X))
max_vif <- max(vif_values)
max_vif_label <- labels[names(which.max(vif_values))]
```

The full model (Table 4) explains `r fmt(summary(m)$r.squared*100,1)`% of the
variation in county premature mortality. Five of six predictors carry the
expected positive
sign and are significant at p < .001: smoking, obesity, uninsured rate, child
poverty, and opioid dispensing. The largest variance inflation factor is
`r fmt(max_vif)` for `r max_vif_label`, well below the common threshold of 5, so
multicollinearity does not indicate a serious threat to these estimates.

```{r standardized-coefficients, fig.height=3.3, fig.cap="Standardized coefficients from the OLS model."}
std_vars <- c("pct_smokers", "pct_obese", "pct_uninsured",
              "grad_rate", "pct_child_poverty", "opioid_rate_2016")
std_df <- data.frame(
  Predictor = labels[std_vars],
  Standardized = as.numeric(co[std_vars, 1]) *
    sapply(analysis[std_vars], sd) / sd(analysis$premature_aamort),
  row.names = NULL
)
std_df <- std_df %>%
  arrange(abs(Standardized)) %>%
  mutate(Predictor = factor(Predictor, levels = Predictor))
ggplot(std_df, aes(Standardized, Predictor)) +
  geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.3) +
  geom_col(fill = "#2C6E91", width = 0.65) +
  geom_text(aes(label = fmt(Standardized, 2)), hjust = -0.15, size = 3) +
  coord_cartesian(xlim = c(0, max(std_df$Standardized) + 0.06)) +
  labs(x = "Standardized coefficient", y = NULL)
```

Because the predictors are measured on different scales, Figure 3 compares
their relative strength using standardised coefficients. This confirms that
child poverty is the strongest predictor, followed by smoking and opioid
dispensing. Obesity and uninsured rate remain positive but smaller, while
graduation rate is close to zero in practical terms.

The exception is **graduation rate**: H4 predicted a negative coefficient, but
the estimate is small and *positive* (b = `r fmt(co["grad_rate",1])`,
p = `r fmt(co["grad_rate",4],3)`), despite a negative bivariate correlation.

```{r nested-models}
m1 <- lm(premature_aamort ~ grad_rate, data = analysis)
g1 <- summary(m1)$coefficients["grad_rate", ]
g2 <- co["grad_rate", ]
nested <- data.frame(
  Model = c("Model 1: graduation rate only", "Model 2: full model (6 predictors)"),
  b = fmt(c(g1[1], g2[1]), 3),
  SE = fmt(c(g1[2], g2[2]), 3),
  p = ifelse(c(g1[4], g2[4]) < .001, "<.001", fmt(c(g1[4], g2[4]), 3)),
  check.names = FALSE
)
kable(nested, booktabs = TRUE, row.names = FALSE,
      align = c("l", "r", "r", "r"),
      caption = "Graduation-rate coefficient (unstandardised) before and after controlling for the other five predictors.")
```

Table 5 makes this explicit: on its own, graduation rate has a negative
coefficient, but it loses that once the other five predictors are controlled.
This sign reversal does not mean education is harmful; it reflects *confounding*
and a *suppression* effect rather than a real property of education. Considered on its own, graduation rate is
negatively correlated with premature mortality
(r = `r fmt(cor(analysis$grad_rate, analysis$premature_aamort))`): counties that
graduate more of their students do tend to have fewer premature deaths. That
bivariate relationship, however, largely reflects the characteristics of
counties with higher graduation rates — such counties also smoke less, are less
obese, and have lower child poverty. Once smoking, obesity, uninsured status, child poverty, and
opioid dispensing are entered as *control variables*, the graduation coefficient
becomes a *conditional association*: the residual relationship that survives after
holding those other determinants constant. Almost none does, which is exactly
what the literature anticipates — education's protective influence operates
largely *through* behaviour and material conditions rather than directly (Puka et
al., 2022). Holding those pathways constant leaves little of the original
negative signal — what remains is a small positive partial association. Because the design is cross-sectional and ecological, this is
offered as an interpretation *consistent with* the data, not as evidence of causal
mediation, and the county-level pattern should not be read as an individual-level
effect (Robinson, 1950).

```{r diagnostics, fig.height=3.2, fig.cap="Residuals vs fitted values for the OLS model."}
diag <- data.frame(fitted = fitted(m), resid = resid(m))
ggplot(diag, aes(fitted, resid)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_hline(yintercept = 0, colour = "#B2182B") +
  geom_smooth(se = FALSE, linewidth = 0.6) +
  labs(x = "Fitted mortality", y = "Residual")
```

The residuals-versus-fitted plot (Figure 4) scatters around zero with no strong
curvature, so the linear specification is reasonable. Their spread widens modestly at higher fitted
values, indicating mild heteroskedasticity; this leaves the coefficient
estimates unbiased but means their standard errors are best read as approximate,
and a robust-standard-error or weighted specification would be a sensible
robustness check.

# Discussion and conclusion

Across more than 2,600 U.S. counties, premature mortality reflects several
social conditions acting together. The
multivariate model supports five of the six hypotheses: behavioural risk
(smoking, obesity), limited healthcare access (uninsured), socioeconomic
disadvantage (child poverty), and a harmful drug environment (opioid
dispensing) are each independently associated with higher premature mortality
(H1, H2, H3, H5, H6), and together they explain about
`r fmt(summary(m)$r.squared*100,0)`% of the cross-county variation. This pattern
is consistent with the social-determinants-of-health framework: place-based
health is shaped jointly by how people behave, how disadvantaged they are,
whether they can reach care, and the local drug environment, rather than by any
single cause. H4 is the exception — graduation rate shows no protective
*independent* effect once the other determinants are controlled — but, as argued
above, this is consistent with education acting *through* behaviour and poverty,
not with education being unimportant.

These findings also illustrate several broader methodological concepts. The graduation
result shows why *confounding* and control variables matter: a bivariate
relationship can change once related social conditions are held constant. The
county-level design also creates an *ecological-inference* problem, so the
coefficients describe places rather than individuals. Finally, the complete-case
sample and mild heteroskedasticity show how sampling and model assumptions shape
how confidently the results can be generalised.

The strongest caveat is design. Because the analysis is *ecological* and
*cross-sectional*, associations measured across counties cannot be transferred
to individuals — a county-level coefficient does not describe any one person's
risk (Robinson, 1950) — and they identify association rather than causation.

Beyond design, the data impose further limits. The complete-case sample reflects
listwise deletion, with missingness concentrated in the graduation-rate and
opioid-dispensing variables, which may bias the sample toward counties with
more complete reporting (typically larger, less rural ones). There is also a
*temporal mismatch*: the opioid measure is from 2016 while the outcome is from
2018, which can be read benignly as a lagged exposure but remains a limitation.
Finally, the model omits plausible *confounders* — county age structure, racial
composition, rurality, healthcare infrastructure, and state-level policy — any of
which could inflate or mask the estimated relationships.

The clearest next step is a spatial or multilevel model that tests the
ecological gap rather than assuming it. State fixed effects would absorb regional
clustering and unobserved state policy; an age-adjusted comparison would separate
demographic structure from health behaviour; and multiple imputation with
sensitivity checks would show how far the complete-case restriction shapes the
results. Even within these limits, the analysis suggests that county premature
mortality is a multi-determined outcome, in which supply-side drug factors and
entrenched socioeconomic disadvantage both warrant policy attention.

# Use of AI

Generative AI tools were used as a support aid: to help structure the report, to
debug and tidy the R code, and to improve wording. All research-design
decisions, variable selection, data handling, statistical interpretation, and
final written content are my own and have been reviewed by me. AI was not used
to fabricate data, results, or citations. References were reviewed for
consistency and relevance before release.

# Bibliography

Aune, D., Sen, A., Prasad, M., Norat, T., Janszky, I., Tonstad, S., Romundstad,
P., & Vatten, L. J. (2016). BMI and all-cause mortality: Systematic review and
non-linear dose-response meta-analysis of 230 cohort studies with 3.74 million
deaths among 30.3 million participants. *BMJ, 353*, i2156.
https://pubmed.ncbi.nlm.nih.gov/27146380/

Centers for Disease Control and Prevention. (2017). *U.S. county opioid
dispensing rates, 2016* [Data set].
https://archive.cdc.gov/www_cdc_gov/drugoverdose/rxrate-maps/county2016.html

Dasgupta, N., Beletsky, L., & Ciccarone, D. (2018). Opioid crisis: No easy fix
to its social and economic determinants. *American Journal of Public Health,
108*(2), 182–186. https://pubmed.ncbi.nlm.nih.gov/29267060/

Flegal, K. M., Kit, B. K., Orpana, H., & Graubard, B. I. (2013). Association of
all-cause mortality with overweight and obesity using standard body mass index
categories: A systematic review and meta-analysis. *JAMA, 309*(1), 71–82.
https://pubmed.ncbi.nlm.nih.gov/23280227/

Freeman, J. D., Kadiyala, S., Bell, J. F., & Martin, D. P. (2008). The causal
effect of health insurance on utilization and outcomes in adults: A systematic
review of US studies. *Medical Care, 46*(10), 1023–1032.
https://pubmed.ncbi.nlm.nih.gov/18815523/

Galobardes, B., Smith, G. D., & Lynch, J. W. (2006). Systematic review of the
influence of childhood socioeconomic circumstances on risk for cardiovascular
disease in adulthood. *Annals of Epidemiology, 16*(2), 91–104.
https://pubmed.ncbi.nlm.nih.gov/16257232/

IHME-CHAIN Collaborators. (2024). Effects of education on adult mortality: A
global systematic review and meta-analysis. *The Lancet Public Health, 9*(3),
e186–e198. https://pubmed.ncbi.nlm.nih.gov/38278172/

Jayedi, A., Khan, T. A., Aune, D., Emadi, A., & Shab-Bidar, S. (2022). Body fat
and risk of all-cause mortality: A systematic review and dose-response
meta-analysis of prospective cohort studies. *International Journal of Obesity,
46*, 1573–1581. https://pubmed.ncbi.nlm.nih.gov/35717418/

Kaplan, R. M., Howard, V. J., Safford, M. M., Howard, G., & Popkin, B. M. (2015).
Educational attainment and longevity: Results from the REGARDS US national
cohort study. *Health Psychology, 34*(4), 398–405.
https://pubmed.ncbi.nlm.nih.gov/25791026/

Lariscy, J. T., Hummer, R. A., & Rogers, R. G. (2018). Cigarette smoking and
all-cause and cause-specific adult mortality in the United States. *Demography,
55*(5), 1855–1885. https://pubmed.ncbi.nlm.nih.gov/30232778/

Monnat, S. M. (2018). Factors associated with county-level differences in U.S.
drug-related mortality rates. *American Journal of Preventive Medicine, 54*(5),
611–619. https://pubmed.ncbi.nlm.nih.gov/29598858/

Mons, U., Müezzinler, A., Gellert, C., Schöttker, B., Abnet, C. C., Bobak, M., de
Groot, L., Freedman, N. D., Jansen, E., Kee, F., Kromhout, D., Kuulasmaa, K.,
Laatikainen, T., O'Doherty, M. G., Bueno-de-Mesquita, B., Orfanos, P., Peters,
A., van der Schouw, Y. T., Wilsgaard, T., . . . Brenner, H. (2015). Impact of
smoking and smoking cessation on cardiovascular events and mortality among older
adults: Meta-analysis of individual participant data from prospective cohort
studies of the CHANCES consortium. *BMJ, 350*, h1551.
https://pubmed.ncbi.nlm.nih.gov/25896935/

Puka, K., Buckley, C., Mulia, N., Lasserre, A. M., Rehm, J., & Probst, C. (2022).
Educational attainment and lifestyle risk factors associated with all-cause
mortality in the US. *JAMA Health Forum, 3*(4), e220401.
https://pubmed.ncbi.nlm.nih.gov/35445213/

Robert Wood Johnson Foundation, & University of Wisconsin Population Health
Institute. (2018). *2018 County Health Rankings national data* [Data set].
https://www.countyhealthrankings.org

Robinson, W. S. (1950). Ecological correlations and the behavior of individuals.
*American Sociological Review, 15*(3), 351–357.
https://doi.org/10.2307/2087176

Stringhini, S., Carmeli, C., Jokela, M., Avendaño, M., Muennig, P., Guida, F.,
Ricceri, F., d'Errico, A., Barros, H., Bochud, M., Chadeau-Hyam, M.,
Clavel-Chapelon, F., Costa, G., Delpierre, C., Fraga, S., Goldberg, M., Giles, G.
G., Krogh, V., Kelly-Irving, M., . . . Zins, M. (2017). Socioeconomic status and
the 25 × 25 risk factors as determinants of premature mortality: A multicohort
study and meta-analysis of 1.7 million men and women. *The Lancet, 389*(10075),
1229–1237. https://pubmed.ncbi.nlm.nih.gov/28159391/

U.S. Department of Health and Human Services. (2014). *The health consequences of
smoking—50 years of progress: A report of the Surgeon General.* Centers for
Disease Control and Prevention. https://www.ncbi.nlm.nih.gov/books/NBK179276/

Wilper, A. P., Woolhandler, S., Lasser, K. E., McCormick, D., Bor, D. H., &
Himmelstein, D. U. (2009). Health insurance and mortality in US adults.
*American Journal of Public Health, 99*(12), 2289–2295.
https://pubmed.ncbi.nlm.nih.gov/19762659/

Wise, P. H. (2016). Child poverty and the promise of human capacity: Childhood as
a foundation for healthy aging. *Academic Pediatrics, 16*(3 Suppl), S37–S45.
https://pubmed.ncbi.nlm.nih.gov/27044700/

Woolhandler, S., & Himmelstein, D. U. (2017). The relationship of health
insurance and mortality: Is lack of insurance deadly? *Annals of Internal
Medicine, 167*(6), 424–431. https://pubmed.ncbi.nlm.nih.gov/28655034/
